---
title: "Crime analysis for problem solving: A first analysis"
description: 
output: 
  learnr::tutorial:
    progressive: true
    css: "css/tutorial_style.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# Load packages
library(learnr)
library(tidyverse)

# Set options
knitr::opts_chunk$set(echo = FALSE, fig.align='center')

# Copy files
if (!dir.exists("css")) dir.create("css")
walk(
  dir("../css/"), 
  ~ file.copy(str_glue("../css/{.}"), str_glue("css/{.}"), overwrite = TRUE)
)
```

## Introduction

As an example of what R can do, in this tutorial we will use R to produce a 
simple crime map showing the locations of several homicide offences. To do this, 
we will skip over lots of the detail of how R works and what choices we should 
make during data analysis. We will return to all these details in future 
sessions.

This tutorial includes short chunks of R code that you can run directly in this
window. We will uses these chunks of code to walk through the code we need to
produce a map. To run the code in each chunk, just click the `Run Code` button
in the top-right corner of each chunk. If you want to re-run a chunk of code,
you can click the `Start Over` button in the top-left corner of each chunk.

When you click `Run Code`, the output of the chunk will appear below it (there
may be a delay of a few seconds for more-complicated chunks).

To try out using a code chunk, click `Run Code` below – you should see the
message 'Hello, world!' printed below the box.

```{r exercise0, exercise=TRUE}
message("Hello, world!")
```

`message()` is an example of a *function* – a piece of R code that does 
something. The thing that the `messages()` function does is simple: it prints a 
message on the screen. You can identify a function in R (and virtually all other
programming languages) because the name of the function is followed by a pair of
parentheses `()` (sometimes the parentheses will have text inside them, but we
will cover that later).

Click `Next Topic` below to continue.



## Loading crime data


### Loading packages

One of the things that makes R so useful for data analysis is that you can 
extend the core features of R by making use of *packages* of functions that have 
been written by R programmers to help you do particular things. To use the 
functions in a package, you have to do two things:

  1. **Install** the package on your machine if it is not already installed --
     once you've installed a package, it stays on your computer and will still
     be there the next time you need it. If a package were a book, installing it
     would be the equivalent of buying it from a bookshop -- except that R 
     packages are free!
  2. **Load** the package, which you need to do at the start of each analysis.
  
For the purpose of this example, the packages you need to make a map have 
already been installed behind the scenes by the software that runs this 
tutorial. To load these packages, you can use the `library()` function. Do this
by clicking `Run Code`. This will produce various messages, all of which you can 
safely ignore for now.



```{r exercise1, exercise=TRUE, exercise.lines=5}
# Road the R packages we need to analyse this data
library("ggmap")
library("sf")
library("tidyverse")
```



### Loading data

The first task in creating any crime map is to obtain the crime and other data
you need. In many cases, preparing the data for analysis and mapping will be a 
substantial task. In this case, we are going to use some pre-prepared crime
data together with a pre-drawn street map (which we will ask R to download
automatically).

The data we will use will be records of homicides in the Downtown neighbourhood
of Atlanta, Georgia, in 2019. We can load the homicide data using the 
`read_csv()` function. Click the `Run Code` button to load the data.

```{r exercise2-setup, include=FALSE}
# load the R packages we need to analyse this data
suppressPackageStartupMessages(library("tidyverse"))
```

```{r exercise2, exercise=TRUE}
# Download the data directly from a website and store it as an object
homicides <- read_csv("https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv")
```

The `read_csv()` function prints a message showing the name of each column in
the data and the type of data (number, text etc) in each column. Again, you can
ignore this message for now.

We have stored the results of the `read_csv()` function in an R *object* called
`homicides`. An object in R is anything that stores any type of data. There are
many types of objects, but for this tutorial we don't need to explore these in
any more detail. All you need to remember for now that: 

  * objects store data and 
  * functions do things.



### Viewing the data

To check the data has been loaded correctly, we can view the it by just typing 
the name of the object in which we stored the data. Click `Run Code` to view the 
data.



```{r exercise3-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("tidyverse"))
# Download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)
```

```{r exercise3, exercise=TRUE}
# display the data
homicides
```



The data contain four columns: a unique identifier for a homicide, a label 
describing when and where that homicide occurred, and the longitude and latitude 
of the homicide location. Depending on the width of your screen, you may need to 
click on the '▸' symbol to view all the columns in the data. We can use this 
data to plot the homicides on a map.



## Processing the data

Before we can plot the data on a map, we have to complete some pre-processing 
steps. Having to process data before being able to analyse or visualise it is 
common in all types of data analysis, but spatial analysis often involves
additional processing that takes account of the special features of spatial
data.



### Converting the data into a spatial format

Two data-processing tasks are needed to produce this map. The first is to 
convert the data into a simple features or SF object, which is a special type of
R object that can be used by functions that process spatial data. We will cover
the details of the `st_as_sf()` function that converts our data into into an SF
object later on. Click `Run Code` to convert the data into SF format.



```{r exercise4-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("tidyverse"))
# Download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)
```

```{r exercise4, exercise=TRUE, exercise.lines=5}
# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(homicides, coords = c("longitude", "latitude"), 
                      crs = 4326, remove = FALSE)
```

When you clicked `Run Code` above, it looked like nothing happened. This is
because the results of the code are stored in the `homicides` object -- when you
store the result produced by an R function in an object, the result is not 
automatically printed. 

Do you remember what R code to use to view the contents of this object? Type 
that code into the box below and click `Run Code` to view the data. If you need 
help, click the `Solution` button.


```{r exercise5-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("tidyverse"))
# download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)
# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(homicides, coords = c("longitude", "latitude"), 
                      crs = 4326, remove = FALSE)
```

```{r exercise5, exercise=TRUE}

```

```{r exercise5-solution}
# To view an object in R, simply type the name of the object. To view the 
# contents of the homicides object, copy the next line into the box below and
# click Run Code.
homicides
```

The data looks identical to before running the function `st_as_sf()`, except 
that there is now an extra column called `geometry` that R understands contains
the location of the each crime on the surface of the earth.



### Defining the area covered by the map

The second data-processing task is to define the area that we want to be covered
by the map. In this case, we will define an area that covers the data and 500
metres in every direction around the data, meaning we can see the area 
surrounding the homicide locations.

Defining the area of the map relative to the data requires us to use several R
functions one after the other, connected using `%>%` (which we will cover 
later). One of the benefits of mapping in R rather than [WYSIWYG](https://en.wikipedia.org/wiki/WYSIWYG)
mapping software is that you can connect all of these commands together and run
them at once, rather than having to open the relevant dialog boxes or clicking
the right button for each command in turn.



```{r exercise6, exercise=TRUE, exercise.setup = "exercise5-setup", exercise.lines=13}
# Define the area included in the map, based on a 500m buffer around the data
map_boundary <- homicides %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_as_sf() %>% 
  st_transform(26967) %>% 
  st_buffer(500 * 3.3) %>% 
  st_transform(4326) %>% 
  st_bbox() %>% 
  as.numeric()

map_boundary
```



This code produces an R object representing the lower-left and upper-right 
corners of our map. Now that we've completed the data processing, we can go on
to produce the map itself.



## Draw the map

We are now ready to produce our map of homicides in downtown Atlanta. So that
people viewing the map will understand where the homicides occurred, we will
plot the homicides on top of a base layer showing streets, parks and other 
geographic features obtained from an 
[online web mapping service](http://maps.stamen.com/). Click `Run Code` to 
create the map.



```{r exercise7-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("ggmap"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("tidyverse"))
# Download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)
# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(homicides, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
# Define the area included in the map, based on a 500m buffer around the data
map_boundary <- homicides %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_as_sf() %>% 
  st_transform(26967) %>% 
  st_buffer(500 * 3.3) %>% 
  st_transform(4326) %>% 
  st_bbox() %>% 
  as.numeric()
```

```{r exercise7, exercise=TRUE, exercise.lines=13, fig.width=7, fig.asp=1}
suppressMessages(get_stamenmap(map_boundary, zoom = 14, maptype = "toner-lite")) %>% 
  ggmap() +
  geom_label(
    aes(longitude, latitude, label = label), data = homicides, size = 2.5,
    lineheight = 1, hjust = 1, nudge_x = -0.0005
  ) +
  geom_point(
    aes(longitude, latitude), data = homicides, colour = "white", 
    fill = "orangered1", size = 4, shape = 21
  ) +
  theme_nothing()
```

You can change the appearance of the map by changing various parts of the code
above and clicking `Run Code` again. For example, you can change the colour of
the points that mark the homicides by changing the code `fill = "orangered1"` to
`fill = "mediumblue"`, or change the base map to a different style by changing
the code `maptype = "toner-lite"` to `maptype = "terrain"`. Each time you change
part of the code, click `Run Code` to see what changes on the map.

Once you have finished experimenting with changing the appearance of the map,
click `Next Topic` below.



## Putting the code together

Now we have walked through the different parts of the code, we can create a map
from scratch in a single block of code. In this example, we will map homicides
from in Glenrose Heights neighbourhood of Atlanta, and a different style of base
map. Since the area covered by the map is derived from the data itself, the
extent of the map will update automatically. Click `Run Code`.


```{r exercise8, exercise=TRUE, exercise.lines=33, fig.width=7, fig.asp=1}
# Load the R packages we need to analyse this data
# suppressPackageStartupMessages() hides the start-up messages we saw earlier
suppressPackageStartupMessages(library("ggmap"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("tidyverse"))

# Download the data directly from a URL and store it as an object
# suppressMessages() hides the column-specification messages we saw earlier
homicides <- suppressMessages(read_csv("https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/glenrose_heights_homicides.csv"))

# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(homicides, coords = c("longitude", "latitude"), 
                      crs = 4326, remove = FALSE)

# Define the area included in the map, based on a 500m buffer around the data
map_boundary <- homicides %>% st_bbox() %>% st_as_sfc() %>% st_as_sf() %>% 
  st_transform(26967) %>% st_buffer(500 * 3.3) %>% st_transform(4326) %>% 
  st_bbox() %>% as.numeric()

# Plot the map
suppressMessages(get_stamenmap(map_boundary, zoom = 14, maptype = "toner")) %>% 
  ggmap() +
  geom_label(
    aes(longitude, latitude, label = label), data = homicides, size = 2.5,
    lineheight = 1, hjust = 0, vjust = 1, nudge_x = 0.0005
  ) +
  geom_point(
    aes(longitude, latitude), data = homicides, colour = "white", 
    fill = "mediumblue", size = 4, shape = 21
  ) +
  theme_nothing()
```



### In summary

This map is effective for showing the locations of just a few crimes, but is
too limited to show more complicated patterns or larger datasets. In the 
following tutorials, we will learn how to do all sorts of crime analysis in R,
including producing other types of crime maps.
